---
title: "statgen_sibs"
author: "Jayden Chrzanowski"
date: '2022-06-29'
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(tidyverse)
library(arsenal)
```
```{r}
# reading phenotype data
phenotype <- read.csv("T2D-GENES_P1_Hispanic_phenotypes.txt", sep="")
head(phenotype)
```
```{r}
# reading genetic data
pcs <- read.csv("all_chr_eigensoft_PCs.csv")
head(pcs)
```
```{r}
phenotype <- subset(phenotype, select = -SMOKE) #removing smoke variable - all NAs
phenotype$HTN <- ifelse(phenotype$SBP>130, 1, ifelse(is.null(phenotype$BPMED), 1, ifelse(phenotype$BPMED==1, 1, 0))) #adding HTN variable
phenotype <- rename(phenotype, "ID" = "IID") # renaming IID in phenotype data to ID to merge with genetic data
phenotype <- merge(phenotype, pcs, by="ID") # merging genetic data by id var
```

```{r}
head(phenotype)
colnames(phenotype)
```
```{r, results='asis'}
# Controls for summary stat tables
my_controls <- tableby.control(
  test = T,
  total = T,
  numeric.test="kwt", cat.test="chisq", #Specify Kruskal-Wallis test for numeric vars and chi square for categorical vars
  numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"), #Display mean, sd, median, q1 and q3, range, and missing values
  cat.stats = c("countpct", "Nmiss2"),
  stats.labels=list(
    meansd = "Mean (SD)",
    medianq1q3 = "Median (Q1, Q3)",
    range = "Min - Max",
    Nmiss2 = "Missing"
  )
)
# Creating labels for variables in table
my_labels <- list(
  DBP = "Diastolic Blood Pressure",
  SBP = "Systolic Blood Pressure",
  AGE = "Age",
  SEX = "Sex",
  BPMED = "Blood Pressure Medication Status",
  HTN = "Hypertension Status"
)

# Creating a list of the PC variables
pc_list <- paste("PC", 1:10, sep="")
htn_fmla <- as.formula(paste("HTN ~ + DBP + SBP + AGE +", paste(pc_list, collapse= "+")))
sex_fmla <- as.formula(paste("SEX ~ + DBP + SBP + AGE +", paste(pc_list, collapse= "+")))
bpmed_fmla <- as.formula(paste("BPMED ~ + DBP + SBP + AGE +", paste(pc_list, collapse= "+")))

htn_table <- tableby(htn_fmla, data = phenotype, control=my_controls) 
summary(htn_table, labelTranslations = my_labels, title = "Summary Statistics for Phenotypes By Hypertension")

sex_table <- tableby(sex_fmla, data = phenotype, control=my_controls) 
summary(sex_table, labelTranslations = my_labels, title = "Summary Statistics for Phenotype Data By Sex")

bpmed_table <- tableby(bpmed_fmla, data = phenotype, control=my_controls) 
summary(bpmed_table, labelTranslations = my_labels, title = "Summary Statistics for Phenotype Data By Blood Pressure Medication Status")

gen_table <- tableby(~DBP + SBP + AGE, data=phenotype, control=my_controls)
summary(gen_table, labelTranslations = my_labels, title="Summary Statistics for Phenotype Data")

summary(freqlist(~SEX, data=phenotype), title="Frequency Table By Sex for Phenotype Data", labelTranslations = my_labels)
summary(freqlist(~BPMED, data=phenotype), title="Frequency Table By Blood Pressure Medication Status for Phenotype Data", labelTranslations = my_labels)
summary(freqlist(~HTN, data=phenotype), title="Frequency Table By Hypertension Status for Phenotype Data", labelTranslations = my_labels)

```
Hypertension has significant associations with DBP, SBP, AGE, PC1, PC2.

Sex has significant associations with DBP, SBP, AGE, and PC2

Blood Pressure Medication Status has significant assocciations with SBP, AGE, and PC4 (most significant PC yet). It makes sense for it not to have an association with DBP since high blood pressure is due to high SBP. It is interesting to note that hypertension does have a significant association with DBP. It appears that those with hypertension have a higher likelihood of having a high DBP.

There were only 3 signficant PCs in this analysis - PC1, PC2, and PC4.

There are many more missing BPMED instances (1536) compared to missing SBP and DBP (92). This does not impact hypertension as it was coded to still code depending on SBP if BPMED was null.

```{r}
boxplot(phenotype$SBP~phenotype$BPMED)
boxplot(phenotype$AGE~phenotype$BPMED)
boxplot(phenotype$PC4~phenotype$BPMED)


bpmed_fmla <- as.formula(paste("BPMED ~ SBP + AGE + SEX + BPMED + HTN + + ", paste(pc_list, collapse="+")))
bpmed_model <- glm(bpmed_fmla,family=binomial(link='logit'), data=phenotype)
#summary(bpmed_model)

bpmed_model_reduc <- glm(BPMED~SBP+AGE+HTN+PC4, family=binomial(link='logit'), data=phenotype)
summary(bpmed_model_reduc)

anova(bpmed_model_reduc, bpmed_model, test="Chisq")
exp(coef(bpmed_model_reduc))

plot(bpmed_model_reduc) # Little shaky on assumptions of linearity, deviation from normality present, potential influential outliers
deviance(bpmed_model_reduc)/df.residual(bpmed_model_reduc) 
```
The reduced model contains the variables SBP, AGE, HTN, and PC4 and appears to be a better fit model than the model with all variables as seen by the p-value of .8925 in the analysis of deviance table.
Each unit increase in systolic blood pressure increases the odds of being on blood pressure medication by 1.02 (2%).
Each year increase in age increases the odds of being on blood pressure medication by 1.05 (5%). 
Having hypertension increases the odds of being on blood pressure medication by 2.71 (171%)

It is important to note that logistic regression assumptions may not be met such as homoskedasticity as seen in the scale-location plot and normality as seen in the normal q-q plot. There are no influential outliers present as seen in the residuals vs leverage plot.
